---
title: "Exploring the UK 2021 Census Data: Gender Identity"
output: html_document
date: "2023-04-19"
---

## LOAD PACKAGES 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(janitor)
library(ggmap)
library(mapview)
library(sf)
library(readxl)
library(janitor)
library(tmap)
library(dplyr)
library(sp)
library(spdep)
library(ggplot2)
library(ggspatial)

```



## LOAD CLEANED AND JO|INED DATA


```{r}
## Read in cleaned and merged data
la_gi <- st_read("/Users/user/Documents/InteractiveGender/Shapefiles/gi_spatial.shp") %>%
  clean_names() %>%
  rename("gi_categories" = g_ctgrs)
head(la_gi)

# This line of code is important. This removes those invalid geometries (those with looped vertices), be patient it takes about 1-2 minutes to run. 
st_is_valid(la_gi, reason = T)
la_gi <- st_make_valid(la_gi)

# And remove NAs, mainly Welsh LAs
sum(is.na(la_gi$observation))
la_gi <- na.omit(la_gi)

## FOR INDIVIDUAL CATS

same <- la_gi %>% filter(gi_categories == "Gender identity the same as sex registered at birth")
diff <- la_gi %>% filter(gi_categories == "Gender identity different from sex registered at birth but no specific identity given")
t_women <- la_gi %>% filter(gi_categories == "Trans woman")
t_man <- la_gi %>% filter(gi_categories == "Trans man")
all_other <- la_gi %>% filter(gi_categories == "All other gender identities")


```



## DATA PRE-PROCESSING 

For those interest in how the data has been cleaned (fixing invalid geometries), joined (to region, urban vs rural, LA population)


### Read in gender identity data

```{r}
gi <- read_csv("Data/gi_renamed.csv") %>% 
  clean_names()

#Delete 'Does not apply' apply category
gi <- gi %>%
  subset(gi_code > 0)

#Add percentages 
gi <- gi %>% 
  group_by(la_code) %>% 
  mutate(total_observation = sum(observation),
         percentage = observation / total_observation * 100)

head(gi$percentage)

```


## Read in LT Local Authority (shapefile)
```{r}
#Load the shapefile
shapefile <- st_read("Shapefiles/LAD_DEC_2022_UK_BFC.shp") %>% clean_names()
#change column name of LSOA code to match the LSOA code column in the SO polygon data
colnames(shapefile)[which(names(shapefile) == "lad22cd")] <- "la_code"

#check crs 
st_crs(shapefile)

#lets transform it to the correct crs
shapefile <- st_transform(shapefile, 4326)

#check the crs again 
st_crs(shapefile)

#Plot the empty shapefile using ggplot or base plot
ggplot() + 
  geom_sf(data = shapefile)

plot(st_geometry(shapefile))

# check class
class(shapefile)
```


## Read in region data

```{r}
region <- read_csv("la_to_region.csv") %>%
  clean_names() %>%
   rename("la_code" = lad22cd, 
         "la_name" = lad22nm,
         "reg_code" = rgn22cd,
         "reg_name" = rgn22nm)
```


## Read in LA population 


```{r}
pop <- read_excel("Data/Census_EW_2021.xlsx", sheet = "P01") %>% 
  clean_names() %>%
  slice(10:382) %>%
  rename("la_code" = p01_census_2021_usual_resident_population_by_sex_local_authorities_in_england_and_wales_note_1,
           "la_name" = x2, 
         "total_pop" = x3,
         "female_pop" = x5,
         "male_pop" = x4)
  
pop$total_pop <- as.numeric(pop$total_pop)
pop$male_pop <- as.numeric(pop$male_pop)
pop$female_pop <- as.numeric(pop$female_pop)

# Remove the regions
remove.list <- paste(c("England", "North West", "North East", "Yorkshire and The Humber", "East Midlands", "West Midlands", "East of England", "London", "South East", "South West", "Wales"), collapse = "|")

df %>% 
  filter(!grepl(remove.list, area_name))



head(pop)
```


## Read in urban/rural data 

```{r}

urb_rur <- read_excel("Data/Rural_Urban_2011.xlsx") %>% clean_names() %>%
  rename("la_code" = lad18cd) %>%
  rename("urb_rur" = broad_ruc11) %>%
  rename("urb_pop" = total_urban_population_2011) %>%
  rename("rur_pop" = total_rural_population_2011)
head(urb_rur)

```


# Join datasets

```{r}
rm(la_gi)

# joining shapefile to GI to make it an sf object
la_gi <- left_join(shapefile, gi, by = "la_code")
class(la_gi)

# joining the sf object to the region table
la_gi <- left_join(la_gi, region, by = "la_code") 
class(la_gi)

# joining the population counts
la_gi <- left_join(la_gi, pop, by = "la_code")

# joining the rban rual 
la_gi <- left_join(la_gi, urb_rur, by = "la_code")


head(la_gi)

```


# Calculate Rates

```{r}
la_gi <- la_gi %>% mutate_at(c('total_pop', 'female_pop', 'male_pop'), as.numeric)

# Create new rate variables from total LA population 
la_gi <- mutate(la_gi, pop_rate = (observation/total_pop)*10000, 
              male_rate = (observation/male_pop)*10000, 
              female_rate = (observation/female_pop)*10000)

head(la_gi)
```




## Saving the spatial object

```{r}
#first remove any irrelevant columns
class(la_gi)
la_gi <- la_gi %>%
  select(,-c(11, 20, 21, 25, 26, 28:30, 32:37))

la_gi <- st_make_valid(la_gi)

st_write(la_gi, "/Users/user/Documents/InteractiveGender/Shapefiles/gi_spatial.shp")



```




## EXPLANTORY ANALYSIS 


## Plot all cats on one map 

```{r}
# Create a color palette for each category
color_pal <- c("Gender identity the same as sex registered at birth" = "blue",
               "Gender identity different from sex registered at birth but no specific identity given" = "green",
               "Trans woman" = "red",
               "Trans man" = "purple",
               "All other gender identities" = "gray")

# Create a ggplot object with the joined data
p <- ggplot() + 
  geom_sf(data = la_gi, aes(fill = gi_categories), color = "gray") + 
  scale_fill_manual(values = color_pal, na.value = "white") +
  labs(title = "Gender Identity in England", fill = "Gender Identity")

# Display the plot
p
```





## Plot seperate cats?

```{r}
## Split genders

tm_shape(same) +
  tm_fill("obsrvtn", style="jenks", title = "same counts") +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = "Gender identity the same as registered at birth", main.title.size = 0.7, 
            legend.position = c ("right", "bottom"), legend.title.size = 0.8)


tm_shape(diff) +
  tm_fill("obsrvtn", style="jenks", title = "same counts") +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = "Gender identity different from sex registered at birth but no specific identity given", main.title.size = 0.7, 
            legend.position = c ("right", "bottom"), legend.title.size = 0.8)


tm_shape(t_women) +
  tm_fill("obsrvtn", style="jenks", title = "same counts") +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = "Trans Women", main.title.size = 0.7, 
            legend.position = c ("right", "bottom"), legend.title.size = 0.8)


tm_shape(t_man) +
  tm_fill("obsrvtn", style="jenks", title = "same counts") +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = "Trans Man", main.title.size = 0.7, 
            legend.position = c ("right", "bottom"), legend.title.size = 0.8)


tm_shape(all_other) +
  tm_fill("obsrvtn", style="jenks", title = "same counts") +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = "All other gender identities", main.title.size = 0.7, 
            legend.position = c ("right", "bottom"), legend.title.size = 0.8)


## to make the map interactive
tmap_mode("view")

#can also add a new line of code i.e. 

```




Leaflet 

```{r}

same$lat <- as.numeric(same$lat)
same$long <- as.numeric(same$long)

# Creating a color palette based on the number range in the total column
pal <- colorNumeric("Greens", domain = same$obsrvtn)

# Getting rid of rows with NA values
# Using the Base R method of filtering subset() because we're dealing with a SpatialPolygonsDataFrame and not a normal data frame, thus filter() wouldn't work
# Setting up the pop up text
popup_sb <- paste0("Total: ", as.character(same$obsrvtn))


leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(53.801277, -1.548567, zoom = 3) %>% 
  addPolygons(data = same , 
              fillColor = ~pal(same$obsrvtn), 
              fillOpacity = 0.7, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              popup = ~popup_sb) %>%
  addLegend(pal = pal, 
            values = same$obsrvtn, 
            position = "bottomright", 
            title = "Gender identity the same")



## with added interactive elements
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(53.801277, -1.548567, zoom = 3) %>% 
  addPolygons(data = same , 
              fillColor = ~pal(same$obsrvtn), 
              fillOpacity = 0.7, 
              weight = 0.2, 
              smoothFactor = 0.2, 
              highlight = highlightOptions(
                  weight = 5,
                  color = "#666",
                  fillOpacity = 0.7,
                   bringToFront = TRUE),
              label=popup_sb,
              labelOptions = labelOptions(
    style = list("font-weight" = "normal", padding = "3px 8px"),
    textsize = "15px",
    direction = "auto")) %>%
  addLegend(pal = pal, 
            values = same$obsrvtn, 
            position = "bottomright", 
            title = "Gender identity the same")

```





## Heat Map 


```{r}
la_gi$broad_ruc11

# create a heatmap using ggplot2
ggplot(la_gi, aes(x = la_code, y = gi_categories, fill = observation)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  facet_wrap(~ broad_ruc11, ncol = 1) +
  labs(title = "Gender Identity Census Data by Local Authority and Location",
       x = "Local Authority Code",
       y = "Gender Identity Category") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

# create an interactive version of the heatmap using plotly
install.packages("plotly")
library(plotly)
ggplotly()


library(tmaptools)
tm_shape(la_gi) + 
  tm_heatmap(col = "gi_categories")



ggplot(la_gi, aes(x = long, y = lat)) + 
  annotation_map_tile() + 
  stat_bin2d(alpha=0.7) + 
  scale_fill_gradientn(colours = c("white","red"), 
                       name = "Frequency")

```










